\section{Galax 1.0.2 (January 2012)}

\section{Galax 1.0 (March 2008)}

Galax version 1.0 implements the XQuery 1.0 Recommendation from
\xqueryrec\ (\xqueryurl) and the XQuery Update Facility 1.0 from
\ultfwd\ (\ultfurl).  It also implements XQueryP, an imperative
scripting language that extends XQuery with updates with mutable
variables, while loops, and sequential expressions (\xqueryp).

\begin{itemize}
\item Galax 1.0 must be built with O'Caml 3.10
\item Implementation of the latest XQuery Update Facility 1.0
(\ultfurl), including static typing.
\item Improved implementation of XQueryP.
\begin{itemize}
      \item Re-implementation of while loops.
      \item Fixed issues with scoping.
\end{itemize}
\item Improved support for modules.
\begin{itemize}
      \item Properly implemented nested imports.
      \item Added support for module interfaces.
\item Support for XQueryX trivial embedding.
\item Bug fixes:
\begin{itemize}
      \item Fixed problems with support for in-scope namespaces.
      \item Fixed problems with checking of cyclic variable declarations.
      \item Fixed several problems with the parser, using wrong lexical states inside constructors.
      \item Small bug fixes in support for the prolog.
\end{itemize}
\item Conformance results on XQTS 1.0.2:
\begin{tabular}{lrrrr}
  Feature & Pass & Fail & Total & Percent\\
  Minimal Conformance      &       14555 & 69 & 14637 & (99.4\%)
  Optional Features \\
    Static Typing Feature   &      46    & 0   & 46 \\
    Full Axis Feature        &     130  & 0   & 130\\
    Module Feature            &    32    & 0   & 32\\
\end{tabular}
\end{itemize}
\end{itemize}

\section{Galax 0.7.2 (February 2007)}

Galax version 0.7.2 is a minor release, and should be considered as a
beta release. Galax 0.7.2 implements the XQuery 1.0 Recommendation
from \xqueryrec.

This is a development release.  Notably static typing and
some of the new compiler optimizations are not fully tested.

\begin{itemize}
\item Compiler:
  \begin{itemize}
  \item We're still working on Join detection...A never-ending saga.
  \end{itemize}

\item Current testing results on XQTS 1.0.2:
\begin{tabular}{lrrrr}
  Feature & Galax Pass & Fail & Total & Percent\\
  Minimal Conformance	 &         14514 & 110 & 14637 & (99.1\%) \\
  Optional Features & \\
    Schema Import Feature&	  0   	& 0   & 174\\
    Schema Validation Feature&	  0   	& 0   & 25\\
    Static Typing Feature	&  46  	& 0   & 46\\
    Full Axis Feature	 	&  130 	& 0   & 130\\
    Module Feature	 	&  0 	& 0   & 32\\
    Trivial XML Embedding Feature &0 	& 0   & 4\\
\end{tabular}
\end{itemize}

\section{Galax 0.6.8 (August 2006)}

Galax version 0.6.8 is a minor release, and should be considered as a
beta release. Galax 0.6.8 implements the XQuery 1.0 candidate
recommendation working drafts from \xqueryversion.

This is a development source-only release.  Notably static typing and
some of the new compiler optimizations are not fully tested.

\begin{itemize}
\item Language: 
  \begin{itemize}
  \item More bug fixes to align with the \xqueryversion CR working
    drafts.
  \item Support for the XQuery! language is fixed (was broken in the
    previous release).
  \item Support for the XML update facility ({\ultfurl}).
  \item Support for the XQueryP extension to XQuery ({\xqueryp}).
  \end{itemize}
\item Compiler:
  \begin{itemize}
  \item Join detection is back! Some bug fixes and improvements to the
    robustness of the optimizer in the presence of rewritings.
  \end{itemize}
\end{itemize}

\section{Galax 0.6.5 (May 2006)}

Galax version 0.6.5 is a major release, and should be considered as an
alpha release. Galax 0.6.5 implements the XQuery 1.0 candidate
recommendation working drafts from \xqueryversion.

This is a development source-only release.  Notably static typing and
some of the new compiler optimizations are not fully tested.

\begin{itemize}
\item Language: 
  \begin{itemize}
  \item Alignment with the \xqueryversion CR working drafts. 
  \item Support for XML updates based on the XQuery! language [Alpha]   
  (See {\xquerybangurl}). 
  \item  Support for both strong and weak typing. [Alpha]. 
  \item Complete implementation of XQuery 1.0 Functions and Operators,
    notably full support for date and time, URI, and QName operations.
  \item Countless bug fixes...
  \end{itemize}
\item Environment:
  \begin{itemize}
  \item New configure script for automated configuration, compilation,
  and installation.  
  \item Galax is now a GODI package({\godiurl}).
  \item Added test harness for the W3C XQuery test-suite.
  Galax passes \textbf{\galaxpassestests} out of \textbf{\galaxtotaltests} tests on version
  \textbf{\xquerytestsuiteversion} of the XQuery 1.0 Test Suite.  The test
  suite contains a total of \textbf{\totaltests} tests. 
  \end{itemize}
\item Compiler:
  \begin{itemize}
  \item Support for a variety of physical algorithms for evaluating
  path expressions, including twig joins, staircase joins, and
  one-pass evaluation over XML token streams. [Alpha]
  \item Known problems:  Support for hash/sort joins is temporarily disabled.
  \end{itemize}
\end{itemize}

\section{Galax 0.5.0 (February 2005)}

Galax version 0.5 is a major release, and should be considered as an
alpha release. Galax 0.5.0 implements the XQuery 1.0 working draft
published in October 2004.

Among the most noticeable changes:

\begin{itemize}
\item Alignment with the XQuery 1.0, October 2004 working drafts.
\item A much faster XML parser, based on Gerd Stolpmann's PXP, fixing
  many XML 1.0 conformance bugs as well.
\item A completely new compiler, including a query optimizer that
  supports join optimizations and should deliver much better
  performances than the previous versions of Galax.
\item ``Document projection'' is finally part of the main release,
  allowing to process queries over large documents. (see the
  glx project command-line tool).
\item Improved support for sorting by document order and duplicate
  removal in the compiler.
\item The Windows port is back, based on the MinGW compilers.
\item New port for MacOS X.
\end{itemize}

Have contributed to this release:
  Mary Fern\'andez, Nicola Onose, Philippe Michiels, Christopher R\'e,
  J\'er\^ome Sim\'eon, Michael Stark.

\section{Galax 0.4.0 (August 2004)}

Galax version 0.4 is a major release, and should be considered as an
alpha release. Galax 0.4.0 implements the latest XQuery 1.0 working
draft published in July 2004. It contains many improvements from the
previous version, as well as new features.

Among the most noticeable improvements and new features: Galax now
comes bundled with Jungle, a simple native XML store. It now supports
XML Schema and "named typing". Finally, it contains some prototype
support for Web services.

Have contributed to this release:
  Mary Fern\'andez, Vladimir Gapeyev, Nicola Onose, Philippe Michiels,
  Doug Petkanics, Christopher R\'e, J\'er\^ome Sim\'eon, Avinash Vyas.

Main changes over the previous version are listed below.

Language changes:
\begin{itemize}
\item Support for the latest XQuery 1.0's specifications (July 2004
  Working Drafts).
\item Support for XML Schema 1.0: schema import, validate, named typing,
  sequence types and type tests.
\item Preliminary support for modules. Import module statements without
  recursion are supported. Details of the semantics will be fixed when
  issues around modules are addressed by the XML Query working group.
\item Support for dates and time.
\item Support for string regular expressions.
\end{itemize}

Environment changes:
\begin{itemize}
\item A new set of command-line tools replace the old ones:
    galax-run:       The XQuery execution engine
    galax-parse:     XML parsing and XML Schema validation
    galax-project:   To apply XML document projection
    galax-mapschema: To map XML Schema to the XQuery type system
    ocaml-galax:     The OCaml interpretor bundled with Galax
\item New command-line tools for Web services:
    galax-mapwsdl:   To map WSDL interfaces to an XQuery module
    xquery2soap:     To deploy an XQuery module as an apache Web
                     service.
\item Revisions to the Caml, C, C++ and Java APIs.
\end{itemize}

Architectural changes:
\begin{itemize}
\item A new extensible data model, making Galax easy to use over 'virtual'
  XML documents.
\item A completely new query compiler and evaluation engine that supports
  an hybrid SAX-tuple-tree algebra. The new compilation infrastructure
  should already show improved performances, although it performs
  little optimizations yet. Expect more work in this area in future
  versions of the system
\end{itemize}

New features:
\begin{itemize}
\item Alpha support for native XML storage with Jungle (on top of
  Sleepycat's BerkeleyDB).
\item Alpha support for calling Web services from within XQuery.
\end{itemize}

Portability:
\begin{itemize}
\item Added Makefile for Mac OSX in config/Makefile.osx.
\item Fixed numerous problems with Win32.
\end{itemize}


\section{Galax 0.3.5 (December 2003)}

This is a bug-fix release:
\begin{itemize}
\item Now compiles with OCaml 3.07.
\item Numerous bug fixes to the XML updates support.
\item Added glx:document-save() function allowing to save the result of a
query (notably useful in the case of an existing document that has
been updated).
\item Fixed bug in the release of Caml values in the C-API.
\item Fixed bug in pretty-printing of function application, and added
pretty-printing for the query prolog.
\item Fixed index bug in fn:substring and fn:translate.
\item Fixed serialization in canonical form.
\item Fixed bug in parsing of PIs in the document root.
\item Fixed bug in validation/casting of atomic values not dealing with
whitespace properly.
\item Fixed bug in key/keyref support introduced with the new API.
\item Fixed bug in parsing of DTD declarations with the PUBLIC keyword.
\end{itemize}


\section{Galax 0.3.1 (June 2003)}

Language extensions:
\begin{itemize}
\item Alpha support for XML updates!
\end{itemize}

Command line:
\begin{itemize}
\item External variables and context items can be bound from the
command-line.
\end{itemize}

API:
\begin{itemize}
\item Brand new, hopefully complete Caml, C, and Java APIs.
  Check them out!
\end{itemize}

Parsing:
\begin{itemize}
\item Switched for good to the SAX parser. glx:document-sax is removed.
\item Complete new support for character encodings. Now detects encoding
in XML declaration properly.  Support for UTF-16.
\end{itemize}


\section{Galax 0.3.0 (January 2003)}

Bugs
\begin{itemize}
\item All reported and fixed bugs are documented in 'Bugs' file.
\end{itemize}

Language:
Numerous changes to align with Nov. 2002 and upcoming Feb 2003 WDs.
\begin{itemize}
\item Grammar alignment:
  'as' SequenceType in type declarations, function signatures, typeswitch
\item Implements element \& attribute constructor semantics of Nov. 2002 WDs.
\item Added positional variables to FLWOR
\item Added support for order-by in FLWOR
\item Implemented complete semantics of path expressions, including
  document order and removing duplicates.
\item Implemented dynamic function dispatch, promotion of arguments to
  arithmetic operators
\item Transitive 'eq' operators 
\end{itemize}

Galax features:
\begin{itemize}
\item Command-line options for monitoring memory and CPU usage.
\end{itemize}

Data model:
\begin{itemize}
\item Updated terminology to align with WDs.
\end{itemize}

Function library:
\begin{itemize}
\item Changed xf: to fn: prefix
\item Added fn:error()
\item Added support for fn:base-uri() and fn:lang()
\item New semantics for fn:data() and fn:boolean()
\end{itemize}


\section{Galax 0.2.0 (October 2002)}

Parsing
\begin{itemize}
\item Fixed very large number of bugs. Support for entities and DTDs is
  still missing.
\item Support for ISO-8859-1 and UTF-8 character encodings.
\item Factorized XML and XQuery parsers. Results in more compact code,
  easier to improve and maintain.
\item Updated XQuery parser to align with latest grammar.
\item Alpha support for SAX-based parsing.
\end{itemize}

Data model:
\begin{itemize}
\item Support for node identity.
\item Fixed support for text nodes.
\end{itemize}

Language:
\begin{itemize}
\item Major revision based on August 16th 2002 working drafts.
\item Full support for XPath expressions. Notably XPath parent, ancestor,
  ancestor-or-self axis. See STATUS for some remaining deviations.
\item Support for node-identity related operations (is, isnot
  distinct-nodes, union, intersect, except, etc.).
\item Support for type promotion in function calls, arithmetics, etc.
\item Fixed many bugs in the semantics, through normalization.
\item Added dynamic semantics for type operations through (simplified)
  form of matching. Still some bugs there.
\end{itemize}

Namespaces:
\begin{itemize}
\item Fixed many bugs in namespace support and printing of namespaces.
\item Implemented support for default function namespace.
\end{itemize}

XML Schema:
\begin{itemize}
\item *Very* alpha support for XML Schema import and validation. Basic
  datatypes are now supported.
\end{itemize}

Type system:
\begin{itemize}
\item Updated type inference with the new language.
\item Made sure all expressions are type checked, necessary for
  optimization.
\item Fixed bugs in typing for typeswitch.
\end{itemize}

Function library:
\begin{itemize}
\item Most of F\&O functions are now implemented. Some limitations apply to
  XML Schema types not yet supported (notably date and time).
\end{itemize}

Optimizer:
\begin{itemize}
\item Support for simple query simplification.
\end{itemize}

Compilation:
\begin{itemize}
\item Removed dependency to the stdlib file.
\item Removed dependency to anything but OCaml compilers and standard unix
  tools.
\item Fixed compilation for both Cygwin and MinGW.
\end{itemize}

Tests:
\begin{itemize}
\item Added large number of regression tests.
\end{itemize}

Tools and interfaces:
\begin{itemize}
\item Removed Java API for now. Will be back soon.
\item Added very limited user-level api (Galapi) in Caml.
\item Changed galax command-lined interpretor. See the new syntax and
  options in ./README
\item Major revision of the pretty-printer for types and XQuery
  expressions. Added support for precedence in both cases.
\end{itemize}

Documentation:
\begin{itemize}
\item Added examples of calls to the Caml and C API's in ./example.
\end{itemize}

